Analysis: Machine Learning - ReqAmt - Outliers

### 2021-12-12 code/analysis_code/analysis-ML-RA-outlier.Rmd
ML update - new models, started outliers

* Null
* Decision Tree
* Random Forest
* Bagged Tree

It compares the  models, and then finally fits the “best” model to the test data. It removes the $4B outlier that corresponds to Hurricane Maria in Puerto Rico in 2017.

@@

* here: for data loading/saving
* tidyverse: for data management
* tidymodels: for data modeling
* skimr: for variable summaries
* broom.mixed: for converting bayesian models to tidy tibbles
* rpart.plot: for visualizing a decision tree
* vip: for variable importance plots
* glmnet: for lasso models
* doParallel: for parallel backend for tuning processes
* ranger: for random forest models
* baguette: for bagged decision tree
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #for data loading/saving
library(tidyverse) #for data management
library(tidymodels) #for data modeling
library(skimr) #for variable summaries
library(broom.mixed) #for converting bayesian models to tidy tibbles
library(rpart.plot) #for visualizing a decision tree
library(vip) #for variable importance plots
library(glmnet) #for lasso models
library(doParallel) #for parallel backend for tuning processes
library(ranger) #for random forest models
library(baguette) #for bagged decision tree

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)
```

@@

* Drop the unused variables
* Perform feature engineering
* Transform Incident Month into ordinal variable
* Remove outlier
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on incidentType
* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
req_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)

#transform incident month into ordinal variable
req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#not enough values per state, so create a variable that is FEMA Regions
req_data$FEMARegion <- ifelse((req_data$state == "CT" |
                                 req_data$state == "ME" |
                                 req_data$state == "MA" |
                                 req_data$state == "NH" |
                                 req_data$state == "RI" |
                                 req_data$state == "VT"),
                              "Region I",
                              ifelse((req_data$state == "NJ" |
                                         req_data$state == "NY" |
                                         req_data$state == "PR" |
                                         req_data$state == "VI"),
                                      "Region II",
                                       ifelse((req_data$state == "DE" |
                                                 req_data$state == "MD" |
                                                 req_data$state == "PA" |
                                                 req_data$state == "VA" |
                                                 req_data$state == "DC" |
                                                 req_data$state == "WV"),
                                              "Region III",
                                               ifelse((req_data$state == "AL" |
                                                         req_data$state == "FL" |
                                                         req_data$state == "GA" |
                                                         req_data$state == "KY" |
                                                         req_data$state == "MS" |
                                                         req_data$state == "NC" |
                                                         req_data$state == "SC" |
                                                         req_data$state == "TN"),
                                                      "Region IV",
                                                      ifelse((req_data$state == "IL" |
                                                               req_data$state == "IN" |
                                                               req_data$state == "MI" |
                                                               req_data$state == "MN" |
                                                               req_data$state == "OH" |
                                                               req_data$state == "WI"),
                                                            "Region V",
                                                             ifelse((req_data$state == "AR" |
                                                                         req_data$state == "LA" |
                                                                         req_data$state == "NM" |
                                                                         req_data$state == "OK" |
                                                                         req_data$state == "TX"),
                                                                      "Region VI",
                                                                       ifelse((req_data$state == "IA" |
                                                                                 req_data$state == "KS" |
                                                                                 req_data$state == "MO" |
                                                                                 req_data$state == "NE"),
                                                                              "Region VII",
                                                                               ifelse((req_data$state == "CO" |
                                                                                         req_data$state == "MT" |
                                                                                         req_data$state == "ND" |
                                                                                         req_data$state == "SD" |
                                                                                         req_data$state == "UT" |
                                                                                         req_data$state == "WY"),
                                                                                      "Region VIII",
                                                                                       ifelse((req_data$state == "AZ" |
                                                                                                 req_data$state == "CA" |
                                                                                                 req_data$state == "HI" |
                                                                                                 req_data$state == "NV"),
                                                                                              "Region IX",
                                                                                               ifelse((req_data$state == "AK" |
                                                                                                         req_data$state == "ID" |
                                                                                                         req_data$state == "OR" |
                                                                                                         req_data$state == "WA"),
                                                                                                      "Region X",
                                                                                                      NA
                                                                                                      )
                                                                                              )
                                                                                      )
                                                                              )
                                                                      )
                                                            )
                                                      )
                                              )
                                     )
                              )

#drop state variable
req_data <- req_data %>%
              dplyr::select(-c(state))

#add to "other" category for incident Types that are infrequent
req_data$incidentType[(req_data$incidentType == "Chemical" |
                        req_data$incidentType == "Coastal Storm" |
                        req_data$incidentType == "Dam/Levee Break" |
                        req_data$incidentType == "Mud/Landslide" |
                        req_data$incidentType == "Volcano" |
                        req_data$incidentType == "Terrorist" |
                        req_data$incidentType == "Snow")] <- "Other"

#remove Hurricane Maria outlier
req_data <-subset(req_data, ReqAmt< 4000000000)

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
req_data_split <- rsample::initial_split(req_data, prop = 2/3,
                                     strata = incidentType)

@@

#5-fold cross validation, repeated 5 times, stratified by incidentType
req_folds <- rsample::vfold_cv(req_train_data,
                               strata = incidentType,
                               v = 5,
                               repeats = 5)

#create recipe that codes categorical variables as dummy variables
RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#RMSE for training data for formatting later
req_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = req_train_data$ReqAmt,
                                              estimate = rep(mean(req_train_data$ReqAmt), nrow(req_train_data))),
                          SE = 0,
                          model = "Null - Train")

@@

#RMSE for testing data for formatting later
req_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = req_test_data$ReqAmt,
                                              estimate = rep(mean(req_test_data$ReqAmt), nrow(req_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#define the tree model
tree_mod <-
  parsnip::decision_tree(cost_complexity = tune(),
                         tree_depth = tune(),
                         min_n = tune()) %>%
  parsnip::set_engine("rpart") %>%
  parsnip::set_mode("regression")

#use the recipe specified earlier
```
---

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = req_folds,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#more detailed plot
DT_detail_plot <- tree_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
DT_detail_plot

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model")
#save manually -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = req_train_data) %>%
                  dplyr::select(.pred, ReqAmt) %>%
                  dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
DT_pred_act

@@

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
                     geom_hline(yintercept = 0, color = "red", lty = 2) +
                     geom_point() +
                     labs(title = "Decision Tree Fit: Residuals",
                          x = "Observation Number",
                          y = "Residual")
DT_resid

@@

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Requested Funds",
                                x = "Predicted Requested Funds ($)",
                                y = "Residual")
DT_pred_resid

@@

The identified best performing tree model has some interesting results. It identifies incident duration, state population, total number of federal agencies, total number of state counties, and individual/households program awardance as significant decision points for predicting requesting FEMA funding. All of these variables are to be expected, but it is interesting that incident type or declaration type are not included as significant. The RMSE is improved from the Null model, but there's still opportunity for improvement. Let's try another model.

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = req_folds,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#plot most important predictors in the model
RF_plot <- vip::vip(x_RF, num_features = 20)
RF_plot

#save manually-- ggsave throws errors

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = req_train_data) %>%
                dplyr::select(.pred, ReqAmt) %>%
                dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point() +
                        labs(title = "RF Fit: Actual vs. Predicted Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
RF_pred_act

@@

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point() +
                            labs(title = "RF Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
RF_resids

@@

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Requested Funds",
                                x = "Fitted Requested Funds ($)",
                                y = "Residual")
RF_pred_resid

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. Interestingly, it includes declaration month of Jan - March as more important than state population. Let's try one more model.

@@

## Bagged Decision Tree Model

@@

#start over with data split to allow for bootstrapping
#set random seed
set.seed(20140102)

#use bootstrap instead of CV
req_data_bootsplit <- req_train_data %>%
                        rsample::bootstraps(times = 10,
                                            strata = "incidentType")

#define model
bag_tree_mod <- baguette::bag_tree(cost_complexity = tune(),
                                   tree_depth = tune(),
                                   min_n = tune()) %>%
                parsnip::set_engine("rpart",
                                    times = 10) %>%
                parsnip::set_mode("regression")

#create recipe that codes categorical variables as dummy variables
BT_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
          step_string2factor(all_nominal())
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
BT_wflow <- workflows::workflow() %>%
               workflows::add_model(bag_tree_mod) %>%
               workflows::add_recipe(BT_rec)
```

@@

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
tree_grid <- dials::grid_regular(cost_complexity(),
                                 tree_depth(),
                                 min_n(),
                                 levels = 5)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
BT_res <- BT_wflow %>%
          tune::tune_grid(resamples = req_data_bootsplit,
                          grid = tree_grid,
                          metrics = metric_set(rmse),
                          control = control_resamples(save_pred = TRUE,
                                                      extract = function(x) extract_model(x)))

#collect metrics
BT_res %>% workflowsets::collect_metrics()

#default visualization
BT_auto_plot <- BT_res %>% autoplot()
BT_auto_plot

#save file
BT_auto_fig_file = here("results","BT-auto-plot.png")
ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)

#more detailed plot
BT_det_plot <- BT_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
BT_det_plot

#save file
BT_det_fig_file = here("results","BT-det-plot.png")
ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
```

@@

### 5. Identify Best Model
```{r}
#select the BT model with the lowest rmse
BT_lowest_rmse <- BT_res %>%
                        tune::select_best("rmse")

#finalize the workflow by using the selected BT model
best_BT_wflow <- BT_wflow %>%
                      tune::finalize_workflow(BT_lowest_rmse)
best_BT_wflow

#one last fit on the training data
best_BT_fit <- best_BT_wflow %>%
                    parsnip::fit(data = req_train_data)
```

@@

### 6. Model evaluation
```{r}
#bag roots function to extract most important variables
bag_roots <-  function(x){
  x %>% 
    dplyr::select(.extracts) %>% 
    unnest(cols = c(.extracts)) %>% 
    dplyr::mutate(models = map(.extracts,
                               ~.x$model_df)) %>% 
    dplyr::select(-.extracts) %>% 
    unnest(cols = c(models)) %>% 
    mutate(root = map_chr(model,
                          ~as.character(.x$fit$frame[1, 1]))) %>%
    dplyr::select(root)  
}

BT_IV <- bag_roots(BT_res) %>% 
            ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
            geom_bar() + 
            coord_flip() + 
            labs(x = "root", y = "count")
BT_IV

#save file
BT_IV_file = here("results","BT-IV-outlier.png")
ggsave(filename = BT_IV_file, plot = BT_IV)

#find predictions and intervals
BT_resid <- best_BT_fit %>%
                  broom.mixed::augment(new_data = req_train_data) %>%
                  dplyr::select(.pred, ReqAmt) %>%
                  dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
BT_pred_act

#save file
BT_pred_act_file = here("results","BT-pred-act.png")
ggsave(filename = BT_pred_act_file, plot = BT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point() +
                      labs(title = "Bagged Tree Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
BT_resids

#save file
BT_resid_file = here("results","BT-resid.png")
ggsave(filename = BT_resid_file, plot = BT_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Requested Funds",
                                x = "Predicted Requested Funds ($)",
                                y = "Residual")
BT_pred_resid

#save file
BT_pred_resid_file = here("results","BT-pred-resid.png")
ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)

#print model performance
#print 10 best performing hyperparameter sets
BT_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
  dplyr::mutate(rmse = round(rmse, 3),
                std_err = round(std_err, 4),
                cost_complexity = scales::scientific(cost_complexity))

#print the best model performance
BT_performance <- BT_res %>% tune::show_best(n = 1)
print(BT_performance)

#compare model performance to null model
BT_RMSE <- BT_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Bagged Tree") %>%
               dplyr::bind_rows(RF_RMSE)
BT_RMSE

#save combined table
model_comp_file = here("results", "ML-model-comp.Rds")
saveRDS(BT_RMSE, file = model_comp_file)
```

In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.

--

>>>>

* Null
* Decision Tree
* Random Forest
* Elastic Net
* LASSO

It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis.

@@

* here: for data loading/saving
* tidyverse: for data management
* tidymodels: for data modeling
* skimr: for variable summaries
* spatialsample: for spatial cross-validation
* broom.mixed: for converting bayesian models to tidy tibbles
* rpart.plot: for visualizing a decision tree
* vip: for variable importance plots
* glmnet: for lasso models
* doParallel: for parallel backend for tuning processes
* ranger: for random forest models
* baguette: for bagged decision tree
* viridis: for a consistent color scheme in figures
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #for data loading/saving
library(tidyverse) #for data management
library(tidymodels) #for data modeling
library(skimr) #for variable summaries
library(spatialsample) #for spatial cross validation
library(broom.mixed) #for converting bayesian models to tidy tibbles
library(rpart.plot) #for visualizing a decision tree
library(vip) #for variable importance plots
library(glmnet) #for lasso models
library(doParallel) #for parallel backend for tuning processes
library(ranger) #for random forest models
library(baguette) #for bagged decision tree
library(viridis) #for a consistent color scheme in figures

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)

#set ggplot theme to classic
ggplot2::theme_set(theme_classic())

#center plot titles
theme_update(plot.title = element_text(hjust = 0.5))
```

@@

* Log transform the outcome
* Drop `ReqAmt`, `disasterNumber`, `declarationTitle`
* Remove the obvious outliers (-log(ReqAmt))
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing
* 5-fold cross validation for the spatial CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#log transform ReqAmt
analysisdata$logReqAmt <- log(analysisdata$ReqAmt)

#drop unnecessary variables
req_data_2 <- analysisdata %>%
              dplyr::select(-c(ReqAmt, disasterNumber, declarationTitle))

#specify all character variables as factors and clean variable names
req_data_1 <- req_data_2 %>% 
                janitor::clean_names() %>%
                dplyr::mutate_if(is.character, factor)

#remove outliers (rows that have a -log(reqamt))
req_data <- req_data_1 %>%
              dplyr::filter(log_req_amt > 0)

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
req_data_split <- rsample::initial_split(req_data, prop = 2/3)

@@

#5-fold spatial cross validation
req_folds_spatial <- spatialsample::spatial_clustering_cv(req_train_data,
                                                          coords = c(latitude, longitude),
                                                          v = 5)

#create recipe that codes categorical variables as dummy variables
RA_rec <- recipes::recipe(log_req_amt ~ ., data = req_train_data) %>%
          recipes::step_unknown(state, new_level = "unknown state") %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#RMSE for training data for formatting later
req_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = req_train_data$log_req_amt,
                                              estimate = rep(mean(req_train_data$log_req_amt), nrow(req_train_data))),
                          SE = 0,
                          model = "Null - Train")

@@

#RMSE for testing data for formatting later
req_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = req_test_data$log_req_amt,
                                              estimate = rep(mean(req_test_data$log_req_amt), nrow(req_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#define the tree model
tree_mod <-
  parsnip::decision_tree(cost_complexity = tune(),
                         tree_depth = tune(),
                         min_n = tune()) %>%
  parsnip::set_engine("rpart") %>%
  parsnip::set_mode("regression")
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = req_folds_spatial,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#more detailed plot
DT_detail_plot <- tree_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free") +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0) +
                             labs(x = "cost complexity",
                                  y = "mean",
                                  color = "Tree Depth")
DT_detail_plot

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model",
                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
                                                     shadow.col = "grey65",
                                                     col = "grey99")
#save manually as "DT-plot.png" -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = req_train_data) %>%
                  dplyr::select(.pred, log_req_amt) %>%
                  dplyr::mutate(.resid = .pred - log_req_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = log_req_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Log(Requested Funds)",
                              x = "Predicted Log(Requested Funds) ($)",
                              y = "Actual Log(Requested Funds) ($)") 
DT_pred_act

@@

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
                     geom_hline(yintercept = 0, color = "red", lty = 2) +
                     geom_point(color = "#3b528b") +
                     labs(title = "Decision Tree Fit: Residuals",
                          x = "Observation Number",
                          y = "Residual")
DT_resid

@@

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Log(Requested Funds)",
                                x = "Predicted Log(Requested Funds) ($)",
                                y = "Residual")
DT_pred_resid

@@

The identified best performing tree model has some interesting results. It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding. New York and November are particularly surprising key predictors. It is interesting that incident type or declaration type are not included as significant. The RMSE actually increases from the Null, which suggests this is not an appropriate model. Moving on.

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = req_folds_spatial,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#plot most important predictors in the model
RF_plot <- vip::vip(x_RF, 
                    num_features = 20,
                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
RF_plot

#save manually as "RF-plot.png" -- ggsave throws errors

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = req_train_data) %>%
                dplyr::select(.pred, log_req_amt) %>%
                dplyr::mutate(.resid = .pred - log_req_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_req_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "RF Fit: Actual vs. Predicted Log(Requested Funds)",
                              x = "Predicted Log(Requested Funds) ($)",
                              y = "Actual Log(Requested Funds) ($)")
RF_pred_act

@@

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point(color = "#3b528b") +
                            labs(title = "RF Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
RF_resids

@@

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Log(Requested Funds)",
                                x = "Fitted Log(Requested Funds) ($)",
                                y = "Residual")
RF_pred_resid

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and whether or not IH was awarded to any part of the state as the five most important predictors. Let's try another model to see if we can find an improvement in the RMSE.

@@

## Elastic Net Model

@@

#define the elastic net model
elastic_mod <- linear_reg(mode = "regression",
                           penalty = tune(),
                           mixture = tune()) %>%
                parsnip::set_engine("glmnet")
```

@@

### 2. Workflow Definition
```{r}
#define elastic net workflow
elastic_wflow <- workflows::workflow() %>%
                 workflows::add_model(elastic_mod) %>%
                 workflows::add_recipe(RA_rec)
```

@@

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
EN_grid <- dials::grid_regular(penalty(),
                               mixture(),
                               levels = 50)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
EN_res <- elastic_wflow %>%
              tune::tune_grid(resamples = req_folds_spatial,
                              grid = EN_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

#look at top 5 EN models
top_EN_models <- EN_res %>%
                    tune::show_best("rmse", n = 5)
top_EN_models

#default visualization
EN_auto_plot <- EN_res %>% autoplot()
EN_auto_plot

#save file
EN_auto_fig_file = here("results","EN-auto-plot.png")
ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
```

@@

### 5. Identify Best Model
```{r}
#select the EN model with the lowest rmse
EN_lowest_rmse <- EN_res %>%
                      tune::select_best("rmse")

#finalize the workflow by using the selected RF model
best_EN_wflow <- elastic_wflow %>%
                      tune::finalize_workflow(EN_lowest_rmse)
best_EN_wflow

#one last fit on the training data
best_EN_fit <- best_EN_wflow %>%
                    parsnip::fit(data = req_train_data)
```

### 6. Model evaluation
```{r}
#extract model from final fit
x_EN <- best_EN_fit$fit$fit$fit

#plot most important predictors in the model
EN_plot <- vip::vip(x_EN, 
                    num_features = 20,
                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
EN_plot

#save manually as "EN-plot.png" -- ggsave throws errors

#find predictions and intervals
EN_resid <- best_EN_fit %>%
                broom.mixed::augment(new_data = req_train_data) %>%
                dplyr::select(.pred, log_req_amt) %>%
                dplyr::mutate(.resid = .pred - log_req_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_req_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "EN Fit: Actual vs. Predicted Log(Requested Funds)",
                              x = "Predicted Log(Requested Funds) ($)",
                              y = "Actual Log(Requested Funds) ($)")
EN_pred_act

#save file
EN_pred_act_file = here("results","EN-pred-act.png")
ggsave(filename = EN_pred_act_file, plot = EN_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point(color = "#3b528b") +
                            labs(title = "EN Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
EN_resids

#save file
EN_resid_file = here("results","EN-resid.png")
ggsave(filename = EN_resid_file, plot = EN_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "EN Fit: Residuals vs Fitted Log(Requested Funds)",
                                x = "Fitted Log(Requested Funds) ($)",
                                y = "Residual")
EN_pred_resid

#save file
EN_pred_resid_file = here("results","EN-pred-resid.png")
ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)

#print the 10 best performing hyperparameter sets
EN_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err)

#print the best model
EN_performance <- EN_res %>% tune::show_best(n = 1)
EN_performance

#compare model performance to null model (and other models)
EN_RMSE <- EN_res %>%
              tune::show_best(n = 1) %>%
              dplyr::transmute(
                rmse = round(mean, 3),
                SE = round(std_err, 4),
                model = "EN") %>%
             dplyr::bind_rows(RF_RMSE)
EN_RMSE
```
The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors. The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors. The EN model has an improved RMSE, but the null model technically still has a lower RMSE.

## LASSO Model

#define the LASSO model
#mixture = 1 identifies the model to be a LASSO model
lasso_mod <- linear_reg(mode = "regression",
                           penalty = tune(),
                           mixture = 1) %>%
                parsnip::set_engine("glmnet")
```
